\textbf{On the Common Logical Structure of Classical and Quantum Mechanics}

### 2021-06-19 Articles/2020-QuantumLogic/Source/quantumlogic.tex
Reviewed through end of Section 3.1

\begin{abstract}
It is generally believed that quantum mechanics entails a rejection of classical logic. The main arguments to support this claim come from the non-commutativity of quantum observables, that allegedly would generate a non-distributive lattice of propositions, and from quantum superpositions, which would entail new rules for quantum disjunctions. In this essay we show that such arguments contain serious flaws: on the one hand, it is possible to keep the classical distributivity law in the context of quantum theory once the actual meaning of quantum propositions is unambiguously defined and properly taken into account. On the other hand, we argue that the existence of quantum superpositions does not lead to the definition of a new rule for logical disjunctions. Furthermore, contrary to the received view in quantum logic, we explain that quantum mechanics can generate a distributive lattice of propositions. Finally, the remarkable formal similarities between the logical structures of classical and quantum mechanics are highlighted and discussed. 
\vspace{4mm}

@@

As soon as Quantum Mechanics (QM) achieved a definite and coherent mathematical formulation, several physicists and philosophers claimed that its formal structure does not conform to the laws of classical propositional calculus, a belief that is widespread still to this day (cf.\ \cite{Giuntini:2002}, \cite{DallaChiara:2004} and \cite{deRonde:2016}). From the thirties, indeed, various modifications of Classical Logic (CL) were advanced in order to faithfully represent the physical content of quantum theory. For instance, in 1931 the Polish philosopher Zygmunt Zawirski proposed to apply \L ukasiewicz's three-valued logic to QM, starting from considerations about the wave-particle duality and Heisenberg's uncertainty principle. In virtue of the latter, moreover, the logical empiricists Moritz Schlick and Philipp Frank claimed that the conjunction of two or more statements may have no meaning in quantum theory, concluding that the classical conjunction is not always valid in the context of quantum physics, as recalled in \cite{Carnap:1966}. In their view, a typical example of invalid conjunction is one attributing a definite value for both position and momentum to a single quantum particle at a given time $t$. Furthermore, in 1933 Fritz Zwicky suggested that QM rejects the law of excluded-middle.\footnote{For historical details on quantum logic see \cite{Jammer:1974}.} However, it is generally recognized that the founding text of Quantum Logic (QL) is Birkhoff and von Neumann's essay ``\emph{The Logic of Quantum Mechanics}'' (\cite{vonNeumann:1936}), where the authors provided the standard account of the propositional calculus obeyed by quantum propositions.\footnote{It is worth noting that von Neumann in his treatise \emph{Mathematische Grundlagen der Quantenmechanik} published in 1932 anticipated that from the algebraic structure of quantum theory it would have been possible to formulate a new propositional calculus. These ideas would have been fully expressed of his successive collaboration with Garrett Birkhoff.}

A few years after the publication of this paper, the research in quantum logic saw a notable development in both the scientific and philosophical communities, opening new research programs searching for the correct logical structures to understand and describe the physics of QM (cf.\ \cite{Reichenbach:1944}, \cite{Mackey:1957}, \cite{Finkelstein:1963}, \cite{Kochen:1965}, \cite{Jauch:1969}, \cite{Giuntini:2002}, \cite{DallaChiara:2004}), and intense metaphysical debates regarding the nature of logic itself and its empirical status (cf.\ notably \cite{Quine:1951}, \cite{Putnam:1968}, \cite{Dummett:1976}, \cite{Hallett:1982}, \cite{Weingartner:2004}, \cite{Bacciagaluppi:2009}). In addition, it is worth noting that research in quantum logic is still very prolific to the present day with many interesting applications in the fields of quantum information, computation and cryptography. Therefore, although QL is no longer viewed as a solution to the foundational problems affecting quantum theory\footnote{This conclusion is now accepted among experts; for details see e.g.\ by \cite{Bacciagaluppi:2009} and \cite{Giuntini:2002}. Referring to this, the latter authors stated explicitly that ``quantum logics are not to be regarded as a kind of ``clue'', capable of solving the main physical and epistemological difficulties of QT [quantum theory]. This was perhaps an illusion of some pioneering workers in quantum logic'' (\cite{Giuntini:2002}, p.\ 225).}, it is an important discipline of contemporary physics. In this paper we will neither take a stance in the dispute concerning the ontological status of logic and its laws, nor we will consider recent developments of QL given that they are not strictly relevant to the point we are going to make. Indeed, the present essay is focused on the logical structures common to both quantum and Classical Mechanics (CM), following the spirit of the former discussion.\footnote{In the present essay we will be concerned only with propositional calculus in the context of non-relativistic QM. First order and higher order logics will not be discussed here as well as the logic of relativistic quantum theories.}

Referring to this, one of us recently faced the question of whether quantum physics implies a revision of classical logic (cf.\ \cite{Oldofredi:2020}). Alternatively stated, it was asked if classical logic must be inevitably abandoned in the quantum realm, or if there are quantum theories compatible with classical propositional calculus. Analyzing Bohmian mechanics, a well-known interpretation of QM, it has been showed that quantum physics does not necessarily involves a departure from classical logic. More precisely, it has been argued that in this interpretation of QM not only the logical connectives retain their classical meanings---in particular the negation and the disjunction do follow classical rules---but also the distributivity law holds. This fact is a consequence of the particle ontology of the theory and its explanation of quantum measurements.

In this paper we will be concerned with a similar issue, since we ask whether standard quantum mechanics itself necessarily involves a new logical structure radically different with respect to classical propositional calculus. In what follows we will answer this question in the negative. More precisely, this essay has a two-fold aim: in the first place, we analyze the main motivations and justifications for the introduction of quantum logic, claiming that such arguments do not really entail the failure of CL in the quantum context because they contain serious flaws. In the second place, we show that the mathematical structure on which QM is built already provides us with a distributive lattice of quantum propositions, contrary to the received view in QL. Such a distributive lattice of quantum statements, moreover, allows us to take into account assertions concerning expectation values, which are generally excluded in the standard approach to quantum logic but are nonetheless experimentally relevant. Finally, we argue that classical probabilistic theory are also equipped with an analogue non-distributive lattice. All these aspects, to our knowledge, are not sufficiently discussed in literature, thus, the present paper can contribute to shed new light on the relation between logic and quantum mechanics.
\vspace{2mm}

The essay is structured as follows: in Section \ref{QL} we review the main arguments that have been advanced in literature to introduce a quantum logical calculus, here we also provide a brief outline of the standard approach to QL (readers familiar with Birkhoff and von Neumann's proposal can skip this part). Section \ref{Motivations} explains why the main examples used to show the failure of classical logic in QM contain important flaws and ambiguities; moreover, we argue that the classical distributive law can be maintained in quantum mechanics, once such ambiguities are removed. In Section \ref{Math}, contrary to the received view in QL, we show that quantum theory already posses a distributive lattice of propositions in the form of the Borel algebra, which includes not only the lattice of standard quantum logic, but also important sentences such as those concerning expectations values. We also show that classical probabilistic theory posses a formally analogue structure. In Section \ref{conc} we discuss the main philosophical implications of the preceding sections and conclude the paper.

@@

The idea is that the logical operations of conjunction, disjunction and negation are replaced respectively by multiplication, addition and complementation among the variables of the theory. This has an interesting physical significance, since the variables associated to magnitudes of physical systems can be added and multiplied together---i.e.\ one can sum and multiply measurement results. Both propositional logic and classical mechanics, thus, generate a complemented Boolean lattice; from this fact it follows that the observable algebra of CM ``is isomorphic to a Boolean algebra of propositions with $(\wedge, \vee, \neg)$'' (\cite{David:2015}, p. 79).
To this regard, Beltrametti claims that 
\begin{quote}
Boolean algebras are algebraic models of classical logic (more specifically of classical propositional calculus) with the algebraic operations of meet ($\wedge$) and join ($\vee$) corresponding to the logical connectives ``and'', ``or'', and the unary relation of orthocomplementation corresponding to the logical negation. The rules and tautologies of classical logic have their counterpart in equations and identities of Boolean algebras (\cite{Beltrametti:2004}, p.\ 341). 
\end{quote}

Therefore, it is possible to formally characterize the state of a physical system via logical propositions which can assume the truth values ``true'' or ``false''---depending whether such statements describe true or false state of affairs concerning the system under consideration (cf.\ \cite{Jaeger:2009}, p. 61)---and to perform logical operations among propositions via the connectives $(\wedge, \vee, \neg)$. It is worth noting that in the context of CM it is always determined whether a system instantiates a given property or not; alternatively stated, every logical proposition about physical systems is either true or false, and hence in CM the principle of semantic bivalence holds (for more details cf.\ \cite{Bub:2007}, p.\ 642, \cite{Giuntini:2002}, p.\ 130).

@@

Contrary to the case of classical mechanics, QM relies on a non-commutative algebraic structure\footnote{For details on the mathematical structure of QM and its physical content the reader may refer e.g.\ to \cite{Sakurai1994}, and \cite{Griffiths:2014}. In this essay, it is assumed that the reader has some familiarity with the standard formalism of QM.}: quantum observables generally do not commute, meaning that for any pair of operators $A,B$ we have $[A,B]=AB-BA\neq0$. From a physical perspective this fact entails that if one performs a measurement of $A$ on a quantum system followed by a measurement of $B$, in general one will obtain different results inverting the order of these observations. Moreover, in virtue of the non-commutative algebra of quantum theory, to measure an operator $C=A+B$ is not typically equivalent to measure $A$ and $B$ independently and then add the respective results, since $A,B$ will be in general incompatible (cf.\ \cite{David:2015}, p.\ 77). This feature of QM constitutes the formal basis to prove the Heisenberg uncertainty relation, a theorem of quantum theory which reflects the operational inability to simultaneously prepare/measure the values of incompatible operators with arbitrary precision. Such a result, in turn, is usually interpreted ontologically in the sense that quantum systems do not instantiate definite properties in non-measurement situations (cf.\ \cite{Sakurai1994}).

@@

Let us consider a $1/2-$spin particle, such as an electron. From QM we know that there are only two possible spin states in which a particle can be found along each axis after a spin measurement, namely either in the up or in the down state. Furthermore, since spin operators along different axes are incompatible and do not commute---i.e.\ $[S_x, S_y]\neq 0, [S_x, S_z]\neq 0, [S_y, S_z]\neq0$---they represent incompatible observables that cannot be simultaneously measured. Let us then consider the following scenario: an electron is prepared with spin up along the $x$ direction and subsequently the spin along the $y$ axis is measured. Let the propositions $p$, $q$ and $r$ be defined as follows:
\begin{description}
	\item $p$: ``the electron has $x$-spin up'', 
	\item $q$: ``the electron has $y$-spin up'',
	\item $r$: ``the electron has $y$-spin down''.
\end{description}
As we said, there are only two states in which a particle can be found along a given axis. Therefore the proposition $q \vee r$ ``must be true'' (\cite{Giuntini:2002}, pp.\ 133-134). However, because the different directions are incompatible, if $p$ is true then both $q$ and $r$ must be false. The result is, as noted before, that the quantum disjunction is true even if neither of its members is true.

@@

Secondly, the authors underlined that in classical and quantum mechanics the states of physical systems are mathematically represented by points in their state spaces---phase space for the classical case, and Hilbert space $\mathcal{H}$ for the quantum case---which provide the maximal information concerning the system. A point in phase space corresponds to the specification of the position and momentum variables of a certain classical system, whereas in QM the points of $\mathcal{H}$ correspond to wave functions. The authors, then, find a connection between subsets of the observation-space of a system and subsets of its Hilbert space, specifying that quantum experimental propositions are mathematically represented by a closed linear subspace of $\mathcal{H}$; this step is crucial in order to obtain the quantum propositional calculus.\ Alternatively, we can say that quantum mechanical operators correspond to propositions with ``yes/no'' (``true/false'') outcome in a logical system\footnote{Cf.\ also \cite{David:2015}, p.\ 78, where we read that ``An orthogonal projector $\textbf{P}$ onto a linear subspace $P\subset\mathcal{H}$ is indeed the operator associated to an observable that can take only the values 1 (and always 1 if the state $\psi\in P$ is in the subspace $P$) or 0 (and always 0 if the state $\psi\in P^{\perp}$ belongs to the orthogonal subspace to $P$).\ Thus we can consider that measuring the observable $\textbf{P}$ is equivalent to perform a test on the system, or to check the validity of a logical proposition $p$ on the system''.} as underlined by Svozi:
\begin{quote}
Any closed linear subspace of---or, equivalently, any projection operator on---a Hilbert space corresponds to an elementary
proposition. The elementary \emph{true--false} proposition can in English be spelled out explicitly as
\begin{quote}
``The physical system has a property corresponding to the associated closed linear subspace'' (\cite{Svozi:1999}, p.\ 1).
\end{quote}
\end{quote}

@@

Now that we have clarified what quantum propositions are, let us introduce Birkhoff and von Neumann's quantum logic. In order to properly define a propositional calculus for QM, one must define the logical operators for conjunction, disjunction and negation and the notion of logical implication. Following \cite{vonNeumann:1936}, the procedure is rather simple: 
\begin{enumerate}
\item The negation of a proposition $p$ is defined by the authors as follows: ``since all operators of quantum mechanics are Hermitian, the mathematical representative of the negative of any experimental proposition is the orthogonal complement of the mathematical representative of the proposition itself'' (\cite{vonNeumann:1936}, pp.\ 826-827). The orthogonal complement $P^{\perp}$ of the subspace $P$ is the set whose elements are all the vectors orthogonal to the elements of $P$. Such an orthogonal complement satisfies the following property: given a subset $P\subset\mathcal{H}$ and a pure state $\psi$, $\psi(P)=1$ iff $\psi(P^{\perp})=0$ and $\psi(P)=0$ iff $\psi(P^{\perp})=1$. As Dalla Chiara and Giuntini underline, ``$\psi$ assigns to an event probability 1 (0, respectively) iff $\psi$ assigns to the orthocomplement of $P$ [notation adapted] probability 0 (1, respectively). As a consequence, one is dealing with an operation that \emph{inverts} the two extreme probability-values, which naturally correspond to the truth-values \emph{truth} and \emph{falsity} (similarly to the classical truth-table of negation)'' (\cite{Giuntini:2002}, p.\ 132).
\item Concerning the conjunction, Birkhoff and von Neumann notice that one can retain the very same set-theoretical interpretation of the classical conjunction in QL, since the intersection of two closed subspaces $P, Q\subset\mathcal{H}$ is still a closed subspace. Thus, one maintains the usual meaning for the conjunction: the pure state $\psi$ verifies $P\cap Q$ iff $\psi$ verifies both $P$ and $Q$. Thus, quantum logic does not introduce a new logical operator for the conjunction.
\item Contrary to the previous case, the logical operator for disjunction cannot be represented by the set-theoretic union, since the set-theoretic union of the two subspaces $P, Q$ will not be in general a subspace, thus, it is not an experimental proposition. Therefore, in QL one introduces the quantum logical disjunction as the the closed span of the subspaces $P, Q$, which is an experimental proposition. Such a statement corresponds to the ``smallest closed subspace containing both $P$ and $Q$'' (\cite{Bacciagaluppi:2009}, p.\ 55). 
\item The logical implication is defined by set theoretic inclusion: given two experimental propositions $p$ and $q$, 
$p$ implies $q$ means that whenever one predicts $p$ with certainty, one can predict also $q$ with certainty, and this is equivalent to stating that $p$ is a subset of $q$. This fact is particularly important since the authors showed that ``it is \emph{algebraically} reasonable to try to correlate physical qualities with subsets of phase-space'' and thus, ``physical qualities attributable to any physical system form a partially ordered system'' (\cite{vonNeumann:1936}, p.\ 828, notation adapted).
\end{enumerate}

@@

However, in the context of QM this move is not allowed: the effect of a measurement on the state of a quantum system, however one may choose to interpret it, cannot be disregarded. A statement about a system before the performance of a certain measurement---i.e.\ about its preparation\footnote{In operational quantum theory the preparations are a set of instructions that an agent has to follow in order to prepare a quantum system in a certain state. We are aware that preparations can be considered special kinds of measurements. For the sake of the argument, however, in what follows we take preparations to be the instructions to arrange a physical system in a given state \emph{before} the measurement of a certain quantity is actually performed.}---is different from a statement regarding the same object when the measurement has been actually carried out, i.e.\ about the obtained measurement outcome. Indeed, it is only through repeated observations that we can infer anything about the state of the system before the measurement. Thus, propositions like ``the electron has $y$-spin up'', ``the velocity of the particle is $v$'', ``the position of the particle is $x$'' are inherently ambiguous as they do not spell out whether they are concerned with preparations or measurement outcomes. Once such vagueness is taken into account and clarified, the purported failure of classical logic disappears.

In order to support this claim, let us redefine the propositions encountered in the previous example as follows:
\begin{description}
	\item $p_i$: ``before the measurement, the electron $x$-spin is up'', 
	\item $p_o$: ``after the measurement, the electron $x$-spin is up'', 
	\item $q_i$: ``before the measurement, the electron $y$-spin is up'',
	\item $q_o$: ``after the measurement, the electron $y$-spin is up'',
	\item $r_i$: ``before the measurement, the electron $y$-spin is down'',
	\item $r_o$: ``after the measurement, the electron $y$-spin is down''.
\end{description}
\noindent Given the setup of the physical situation at hand, after the measurement of $S_y$ either $q_o$ or $r_o$ must be true, which means $q_o \vee r_o = \top$. This reflects the simple fact that \emph{after} a measurement of spin along the $y$ axis the state of the quantum particle must be one of the two indicated in either $q_o$ or $r_o$. Nonetheless, it is worth noting that $q_i$ and $r_i$ are \emph{not} the same statements as $q_o$ and $r_o$---i.e.\ $q_i$ and $r_i$ are not equivalent to $q_o$ and $r_o$---therefore, it is not necessarily true that $q_o \vee r_o = q_i \vee r_i$. This unwarranted equivalence is what causes the problem, to make evident why it is so, let us add a third step to our example.

>>>>

\begin{abstract}
It is generally believed that quantum mechanics entails a rejection of classical logic. The main arguments to support this claim come from the non-commutativity of quantum observables, which allegedly would generate a non-distributive lattice of propositions, and from quantum superpositions, which would entail new rules for quantum disjunctions. In this essay we show that such arguments contain serious flaws.  In particular, it is possible to keep the classical distributivity law in the context of quantum theory once the actual meaning of quantum propositions is unambiguously defined and properly taken into account. We moreover argue that the existence of quantum superpositions does not lead to the definition of a new rule for logical disjunctions. Furthermore, contrary to the received view in quantum logic, we explain that quantum mechanics can generate a distributive lattice of propositions. Finally, the remarkable formal similarities between the logical structures of classical and quantum mechanics are highlighted and discussed. 
\vspace{4mm}

@@

As soon as Quantum Mechanics (QM) achieved a definite and coherent mathematical formulation, several physicists and philosophers claimed that its formal structure does not conform to the laws of classical propositional calculus, a belief that is widespread still to this day (cf.\ \cite{Giuntini:2002}, \cite{DallaChiara:2004} and \cite{deRonde:2016}). From the thirties, indeed, various modifications of Classical Logic (CL) were advanced in order to faithfully represent the physical content of quantum theory. For instance, in 1931 the Polish philosopher Zygmunt Zawirski proposed to apply \L ukasiewicz's three-valued logic to QM, starting from considerations about the wave-particle duality and Heisenberg's uncertainty principle. In virtue of the latter, moreover, the logical empiricists Moritz Schlick and Philipp Frank claimed that the conjunction of two or more statements may have no meaning in quantum theory, concluding that the classical conjunction is not always valid in the context of quantum physics, as recalled in \cite{Carnap:1966}. In their view, a typical example of invalid conjunction is one attributing a definite value for both position and momentum to a single quantum particle at a given time $t$. Furthermore, in 1933 Fritz Zwicky suggested that QM rejects the law of excluded-middle.\footnote{For historical details on quantum logic see \cite{Jammer:1974}.} However, it is generally recognized that the founding text of Quantum Logic (QL) is Birkhoff and von Neumann's essay ``\emph{The Logic of Quantum Mechanics}'' (\cite{vonNeumann:1936}), where the authors provided the standard account of the propositional calculus obeyed by quantum propositions.\footnote{It is worth noting that von Neumann in his treatise \emph{Mathematische Grundlagen der Quantenmechanik} published in 1932 anticipated that from the algebraic structure of quantum theory it would have been possible to formulate a new propositional calculus. These ideas would be fully expressed in his successive collaboration with Garrett Birkhoff.}

A few years after the publication of this paper, the research in quantum logic saw a notable development in both the scientific and philosophical communities, opening new research programs searching for the correct logical structures to understand and describe the physics of QM (cf.\ \cite{Reichenbach:1944}, \cite{Mackey:1957}, \cite{Finkelstein:1963}, \cite{Kochen:1965}, \cite{Jauch:1969}, \cite{Giuntini:2002}, \cite{DallaChiara:2004}), and intense metaphysical debates regarding the nature of logic itself and its empirical status (cf.\ notably \cite{Quine:1951}, \cite{Putnam:1968}, \cite{Dummett:1976}, \cite{Hallett:1982}, \cite{Weingartner:2004}, \cite{Bacciagaluppi:2009}). In addition, it is worth noting that research in quantum logic is still very prolific to the present day with many interesting applications in the fields of quantum information, computation and cryptography. Therefore, although QL is no longer viewed as a solution to the foundational problems affecting quantum theory\footnote{This conclusion is now accepted among experts; for details see e.g.\ \cite{Bacciagaluppi:2009} and \cite{Giuntini:2002}. Referring to this, the latter authors stated explicitly that ``quantum logics are not to be regarded as a kind of ``clue'', capable of solving the main physical and epistemological difficulties of QT [quantum theory]. This was perhaps an illusion of some pioneering workers in quantum logic'' (\cite{Giuntini:2002}, p.\ 225).}, it remains an important discipline of contemporary physics. In this paper we will neither take a stance in the dispute concerning the ontological status of logic and its laws, nor we will consider recent developments of QL given that they are not strictly relevant to the point we are going to make. Indeed, the present essay is focused on the logical structures common to both quantum and classical mechanics, following the spirit of the former discussion.\footnote{In the present essay we will be concerned only with propositional calculus in the context of non-relativistic QM. First-order and higher-order logics will not be discussed here, nor the logic of relativistic quantum theories.}

Referring to this, one of us recently faced the question of whether quantum physics implies a revision of classical logic (cf.\ \cite{Oldofredi:2020}). Alternatively stated, it was asked if classical logic must be inevitably abandoned in the quantum realm, or if there are quantum theories compatible with classical propositional calculus. Analyzing Bohmian mechanics, a well-known interpretation of QM, it has been shown that quantum physics does not necessarily involve a departure from classical logic. More precisely, it has been argued that in this interpretation of QM not only the logical connectives retain their classical meanings---in particular the negation and the disjunction do follow classical rules---but also the distributivity law holds. This fact is a consequence of the particle ontology of the theory and its explanation of quantum measurements.

In this paper we will be concerned with a similar issue, since we ask whether standard quantum mechanics itself necessarily involves a new logical structure radically different with respect to classical propositional calculus. In what follows we will answer this question in the negative. More precisely, this essay has a two-fold aim: in the first place, we analyze the main motivations and justifications for the introduction of quantum logic, claiming that such arguments do not really entail the failure of CL in the quantum context because they contain serious flaws. In the second place, we show that the mathematical structure on which QM is built already provides us with a distributive lattice of quantum propositions, contrary to the received view in QL. Such a distributive lattice of quantum statements, moreover, allows us to take into account assertions concerning expectation values, which are generally excluded in the standard approach to quantum logic but are nonetheless experimentally relevant. Finally, we argue that classical probabilistic theory is also equipped with an analogue non-distributive lattice. All these aspects, to our knowledge, are not sufficiently discussed in literature; thus, the present paper can help to shed new light on the relation between logic and quantum mechanics.
\vspace{2mm}

The essay is structured as follows: in Section \ref{QL} we review the main arguments that have been advanced in the literature to introduce a quantum logical calculus; here we also provide a brief outline of the standard approach to QL (readers familiar with Birkhoff and von Neumann's proposal can skip this part). Section \ref{Motivations} explains why the main examples used to show the failure of classical logic in QM contain important flaws and ambiguities; moreover, we argue that the classical distributive law can be maintained in quantum mechanics, once such ambiguities are removed. In Section \ref{Math}, contrary to the received view in QL, we show that quantum theory already possesses a distributive lattice of propositions in the form of the Borel algebra, which includes not only the lattice of standard quantum logic, but also important sentences such as those concerning expectations values. We also show that classical probabilistic theory possesses a formally analogue structure. In Section \ref{conc} we discuss the main philosophical implications of the preceding sections and conclude the paper.

@@

The idea is that the logical operations of conjunction, disjunction and negation are replaced respectively by multiplication, addition and complementation among the variables of the theory. This has an interesting physical significance, since the variables associated to magnitudes of physical systems can be added and multiplied together---i.e.\ one can sum and multiply measurement results. Both propositional logic and classical mechanics, thus, generate a complemented Boolean lattice; from this fact it follows that the observable algebra of CM ``is isomorphic to a Boolean algebra of propositions with $(\wedge, \vee, \neg)$'' (\cite{David:2015}, p. 79).
In this regard, Beltrametti claims that 
\begin{quote}
Boolean algebras are algebraic models of classical logic (more specifically of classical propositional calculus) with the algebraic operations of meet ($\wedge$) and join ($\vee$) corresponding to the logical connectives ``and'', ``or'', and the unary relation of orthocomplementation corresponding to the logical negation. The rules and tautologies of classical logic have their counterpart in equations and identities of Boolean algebras (\cite{Beltrametti:2004}, p.\ 341). 
\end{quote}

Therefore, it is possible to formally characterize the state of a physical system via logical propositions which can assume the truth values ``true'' or ``false''---depending whether such statements describe a true or false state of affairs concerning the system under consideration (cf.\ \cite{Jaeger:2009}, p. 61)---and to perform logical operations among propositions via the connectives $(\wedge, \vee, \neg)$. It is worth noting that in the context of CM it is always determined whether a system instantiates a given property or not; alternatively stated, every logical proposition about physical systems is either true or false, and hence in CM the principle of semantic bivalence holds (for more details cf.\ \cite{Bub:2007}, p.\ 642, \cite{Giuntini:2002}, p.\ 130).

@@

Contrary to the case of classical mechanics, QM relies on a non-commutative algebraic structure\footnote{For details on the mathematical structure of QM and its physical content the reader may refer e.g.\ to \cite{Sakurai1994}, and \cite{Griffiths:2014}. In this essay, it is assumed that the reader has some familiarity with the standard formalism of QM.}: quantum observables generally do not commute, meaning that for any pair of operators $A,B$ we have $[A,B]=AB-BA\neq0$. From a physical perspective this fact entails that if one performs a measurement of $A$ on a quantum system followed by a measurement of $B$, in general one will obtain different results inverting the order of these observations. Moreover, in virtue of the non-commutative algebra of quantum theory, measuring an operator $C=A+B$ is not typically equivalent to measuring $A$ and $B$ independently and then adding the respective results, since $A,B$ will be in general incompatible (cf.\ \cite{David:2015}, p.\ 77). This feature of QM constitutes the formal basis to prove the Heisenberg uncertainty relation, a theorem of quantum theory which reflects the operational inability to simultaneously prepare/measure the values of incompatible operators with arbitrary precision. Such a result, in turn, is usually interpreted ontologically in the sense that quantum systems do not instantiate definite properties in non-measurement situations (cf.\ \cite{Sakurai1994}).

@@

Let us consider a spin-$1/2$ particle, such as an electron. From QM we know that there are only two possible spin states in which a particle can be found along each axis after a spin measurement, namely either in the up or in the down state. Furthermore, since spin operators along different axes are incompatible and do not commute---i.e.\ $[S_x, S_y]\neq 0, [S_x, S_z]\neq 0, [S_y, S_z]\neq0$---they represent incompatible observables that cannot be simultaneously measured. Let us then consider the following scenario: an electron is prepared with spin up along the $x$ direction and subsequently the spin along the $y$ axis is measured. Let the propositions $p$, $q$ and $r$ be defined as follows:
\begin{description}
	\item $p$: ``the electron has $x$-spin up'', 
	\item $q$: ``the electron has $y$-spin up'',
	\item $r$: ``the electron has $y$-spin down''.
\end{description}
As we said, there are only two states in which a particle can be found along a given axis. Therefore the proposition $q \vee r$ ``must be true'' (\cite{Giuntini:2002}, pp.\ 133-134). However, because the different directions are incompatible, if $p$ is true then both $q$ and $r$ must be false. The result is, as noted before, that the quantum disjunction is true even if neither of its members is true.

@@

Secondly, the authors underlined that in classical and quantum mechanics the states of physical systems are mathematically represented by points in their state spaces---phase space for the classical case, and Hilbert space $\mathcal{H}$ for the quantum case---which provide the maximal information concerning the system. A point in phase space corresponds to the specification of the position and momentum variables of a certain classical system, whereas in QM the points of $\mathcal{H}$ correspond to wave functions. The authors, then, find a connection between subsets of the observation-space of a system and subsets of its Hilbert space, specifying that quantum experimental propositions are mathematically represented by a closed linear subspace of $\mathcal{H}$; this step is crucial in order to obtain the quantum propositional calculus.\ Alternatively, we can say that quantum mechanical operators correspond to propositions with ``yes/no'' (``true/false'') outcomes in a logical system\footnote{Cf.\ also \cite{David:2015}, p.\ 78, where we read that ``An orthogonal projector $\textbf{P}$ onto a linear subspace $P\subset\mathcal{H}$ is indeed the operator associated to an observable that can take only the values 1 (and always 1 if the state $\psi\in P$ is in the subspace $P$) or 0 (and always 0 if the state $\psi\in P^{\perp}$ belongs to the orthogonal subspace to $P$).\ Thus we can consider that measuring the observable $\textbf{P}$ is equivalent to perform a test on the system, or to check the validity of a logical proposition $p$ on the system''.} as underlined by Svozi:
\begin{quote}
Any closed linear subspace of---or, equivalently, any projection operator on---a Hilbert space corresponds to an elementary
proposition. The elementary \emph{true--false} proposition can in English be spelled out explicitly as
\begin{quote}
``The physical system has a property corresponding to the associated closed linear subspace'' (\cite{Svozi:1999}, p.\ 1).
\end{quote}
\end{quote}

@@

Now that we have clarified what quantum propositions are, let us introduce Birkhoff and von Neumann's quantum logic. In order to properly define a propositional calculus for QM, one must define the logical operators for conjunction, disjunction and negation and the notion of logical implication. Following \cite{vonNeumann:1936}, the procedure is rather simple: 
\begin{enumerate}
\item The negation of a proposition $p$ is defined by the authors as follows: ``since all operators of quantum mechanics are Hermitian, the mathematical representative of the negative of any experimental proposition is the orthogonal complement of the mathematical representative of the proposition itself'' (\cite{vonNeumann:1936}, pp.\ 826-827). The orthogonal complement $P^{\perp}$ of the subspace $P$ is the set whose elements are all the vectors orthogonal to the elements of $P$. Such an orthogonal complement satisfies the following property: given a subset $P\subset\mathcal{H}$ and a pure state $\psi$, $\psi(P)=1$ iff $\psi(P^{\perp})=0$ and $\psi(P)=0$ iff $\psi(P^{\perp})=1$. As Dalla Chiara and Giuntini underline, ``$\psi$ assigns to an event probability 1 (0, respectively) iff $\psi$ assigns to the orthocomplement of $P$ [notation adapted] probability 0 (1, respectively). As a consequence, one is dealing with an operation that \emph{inverts} the two extreme probability-values, which naturally correspond to the truth-values \emph{truth} and \emph{falsity} (similarly to the classical truth-table of negation)'' (\cite{Giuntini:2002}, p.\ 132).
\item Concerning the conjunction, Birkhoff and von Neumann note that one can retain the very same set-theoretical interpretation of the classical conjunction in QL, since the intersection of two closed subspaces $P, Q\subset\mathcal{H}$ is still a closed subspace. Thus, one maintains the usual meaning for the conjunction: the pure state $\psi$ verifies $P\cap Q$ iff $\psi$ verifies both $P$ and $Q$. Thus, quantum logic does not introduce a new logical operator for the conjunction.
\item Contrary to the previous case, the logical operator for disjunction cannot be represented by the set-theoretic union, since the set-theoretic union of the two subspaces $P, Q$ will not be in general a subspace; thus, it is not an experimental proposition. Therefore, in QL one introduces the quantum logical disjunction as the the closed span of the subspaces $P, Q$, which is an experimental proposition. Such a statement corresponds to the ``smallest closed subspace containing both $P$ and $Q$'' (\cite{Bacciagaluppi:2009}, p.\ 55). 
\item The logical implication is defined by set theoretic inclusion: given two experimental propositions $p$ and $q$, 
$p$ implies $q$ means that whenever one predicts $p$ with certainty, one can predict also $q$ with certainty, and this is equivalent to stating that $p$ is a subset of $q$. This fact is particularly important since the authors showed that ``it is \emph{algebraically} reasonable to try to correlate physical qualities with subsets of phase-space'' and thus, ``physical qualities attributable to any physical system form a partially ordered system'' (\cite{vonNeumann:1936}, p.\ 828, notation adapted).
\end{enumerate}

@@

However, in the context of QM this move is not allowed: the effect of a measurement on the state of a quantum system, however one may choose to interpret it, cannot be disregarded. A statement about a system before the performance of a certain measurement---i.e.\ about its preparation\footnote{In operational quantum theory the preparations are a set of instructions that an agent has to follow in order to prepare a quantum system in a certain state. We are aware that preparations can be considered special kinds of measurements. For the sake of the argument, however, in what follows we take preparations to be the instructions to arrange a physical system in a given state \emph{before} the measurement of a certain quantity is actually performed.}---is different from a statement regarding the same object when the measurement has been actually carried out, i.e.\ about the obtained measurement outcome. Indeed, it is only through repeated observations on identically prepared systems that we can infer anything about the state of the system before the measurement. Thus, propositions like ``the electron has $y$-spin up'', ``the velocity of the particle is $v$'', ``the position of the particle is $x$'' are inherently ambiguous as they do not spell out whether they are concerned with preparations or measurement outcomes. Once such vagueness is taken into account and clarified, the purported failure of classical logic disappears.

In order to support this claim, let us redefine the propositions encountered in the previous example as follows:
\begin{description}
	\item $p_i$: ``before the measurement, the electron $x$-spin is up'', 
	\item $p_o$: ``after the measurement, the electron $x$-spin is up'', 
	\item $q_i$: ``before the measurement, the electron $y$-spin is up'',
	\item $q_o$: ``after the measurement, the electron $y$-spin is up'',
	\item $r_i$: ``before the measurement, the electron $y$-spin is down'',
	\item $r_o$: ``after the measurement, the electron $y$-spin is down''.
\end{description}
\noindent Given the setup of the physical situation at hand, after the measurement of $S_y$ either $q_o$ or $r_o$ must be true, which means $q_o \vee r_o = \top$. This reflects the simple fact that \emph{after} a measurement of spin along the $y$ axis the state of the quantum particle must be one of the two indicated in either $q_o$ or $r_o$. Nonetheless, it is worth noting that $q_i$ and $r_i$ are \emph{not} the same statements as $q_o$ and $r_o$---i.e.\ $q_i$ and $r_i$ are not equivalent to $q_o$ and $r_o$---therefore, it is not necessarily true that $q_o \vee r_o = q_i \vee r_i$. This unwarranted equivalence is what causes the problem. To make evident why it is so, let us add a third step to our example.
